---
title: "HPNE EV and KRAS - 4i Analysis"
author: "Gaith Droby"
date: "2024-04-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggridges)
```

#### Data Import from standardized 4i output
```{r}
# List all .csv files in the current directory
std_data <- list.files(path = ".", pattern = '*.csv')

# Define sample names
sampleName <- c('EV', 'EV 200', 'KRAS', 'KRAS 200')

featureName <- c('1.DNA', 'EdU', 'CDK2', 'pRB', 'Rb', 'pRB.RB', 'cycA2', 'pp53', 'pH2AX', 'pp27', 'p21', 'pp21')

# Use lapply to apply the function to each element in the list
data_frames <- lapply(std_data, read.csv)

# Assign sample names to each data frame
names(data_frames) <- sampleName

# Use lapply to apply the function to each element in the list
data_frames <- lapply(data_frames, function(df) {
    colnames(df) <- featureName
    return(df)
})
```

#### Remove outliers based on DNA content (>4N DNA)
```{r}
# Use lapply to apply the function to each element in the list
data_frames <- lapply(data_frames, function(df) {
    df <- df[df$'1.DNA' <= 5, ]
    df <- df[apply(df, 1, function(x) all(x <= 7.5)),]
    return(df)
})
```

#### Label G1, S, and G2/M phases of each condition (based on DNA content)
```{r}
# Use lapply to apply the function to each element in the list
g1_data_frames <- lapply(data_frames, function(df) {
    df <- df[df$'1.DNA' <= 0, ]
    return(df)
})

s_data_frames <- lapply(data_frames, function(df) {
    df <- df[df$'1.DNA' >= 0, ]
    df <- df[df$'1.DNA' <= 1.25, ]
    return(df)
})

g2_data_frames <- lapply(data_frames, function(df) {
    df <- df[df$'1.DNA' >= 1.25, ]
    return(df)
})
```


#### Plot features in violin plots 
```{r}
plot_violin <- function(df, name) {
  # Add a column with the sample name to the data frame
  df$SampleName <- name
  
  # Reshape the data frame to long format
  long_df <- gather(df, key = "Feature", value = "Measurement", -SampleName)
  
  # Create the violin plot with different colors for each Feature
  p <- ggplot(long_df, aes(x = Feature, y = Measurement, fill = Feature)) + # Change here
    geom_violin(trim = FALSE) +
    scale_fill_viridis_d() + # Use viridis color palette for the fill
    labs(title = paste("Violin Plot for", name)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) # Rotate x labels for readability
  
  # Save the plot to the working directory
  ggsave(filename = paste("smpl_vioPlt_", name, ".pdf", sep = ""), plot = p, width = 10, height = 6)
  
  return(p)
}

# Loop through the list of data frames and plot each one with its sample name
plots <- mapply(plot_violin, data_frames, names(data_frames), SIMPLIFY = FALSE)

# print them one by one
for (p in plots) {
  print(p)
}
```

#### Plot features in violin plots 
```{r echo=FALSE}
plot_violin <- function(df, name) {
  # Add a column with the sample name to the data frame
  df$SampleName <- name
  
  # Reshape the data frame to long format
  long_df <- gather(df, key = "Feature", value = "Measurement", -SampleName)
  
  # Create the violin plot with different colors for each Feature
  p <- ggplot(long_df, aes(x = Feature, y = Measurement, fill = Feature)) + # Change here
    geom_violin(trim = FALSE) +
    scale_fill_viridis_d() + # Use viridis color palette for the fill
    labs(title = paste("Violin Plot for", name)) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) # Rotate x labels for readability
  
  # Save the plot to the working directory
  #ggsave(filename = paste("smpl_vioPlt_", name, ".pdf", sep = ""), plot = p, width = 10, height = 6)
  
  return(p)
}

# Loop through the list of data frames and plot each one with its sample name
plots <- mapply(plot_violin, s_data_frames, names(s_data_frames), SIMPLIFY = FALSE)

# print them one by one
for (p in plots) {
  print(p)
}

```

# Compare the distribution of each feature in the EV and KRAS cells
```{r}
# Loop through each feature
for (feature in featureName) {
  # Combine the current feature column from each data frame into one data frame
  combined_df <- do.call(rbind, lapply(seq_along(data_frames), function(i) {
    df <- data_frames[[i]][, feature, drop = FALSE] # Select the current feature column
    df$SampleName <- sampleName[i] # Add the sample name
    df$FeatureName <- feature # Add the feature name
    return(df)
  }))
  
  # Rename the feature column to a generic name for plotting
  names(combined_df)[1] <- "Measurement"
  
  # Create the violin plot
  p <- ggplot(combined_df, aes(x = SampleName, y = Measurement, fill = SampleName)) +
    geom_violin(trim = FALSE) +
    scale_fill_brewer(palette = "Set3") + # Use a color palette for the fill
    labs(title = paste("Violin Plot for", feature)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x labels for readability
  
  # Save the plot to the working directory with a unique name for each feature
  ggsave(paste("comp_vioPlt_", feature, ".pdf", sep = ""), plot = p, width = 10, height = 6)
  
    print(p)
}
```


# Compare the G1-phase distribution of each feature in the EV and KRAS cells
```{r eval=FALSE, include=FALSE}
# Loop through each feature
for (feature in featureName) {
  # Combine the current feature column from each data frame into one data frame
  combined_df <- do.call(rbind, lapply(seq_along(g1_data_frames), function(i) {
    df <- g1_data_frames[[i]][, feature, drop = FALSE] # Select the current feature column
    df$SampleName <- sampleName[i] # Add the sample name
    df$FeatureName <- feature # Add the feature name
    return(df)
  }))
  
  # Rename the feature column to a generic name for plotting
  names(combined_df)[1] <- "Signal"
  
  # Create the violin plot
  p <- ggplot(combined_df, aes(x = SampleName, y = Signal, fill = SampleName)) +
    geom_violin(trim = FALSE) +
    scale_fill_brewer(palette = "Set3") + # Use a color palette for the fill
    labs(title = paste("G1-phase violin Plot for", feature)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x labels for readability
  
  # Save the plot to the working directory with a unique name for each feature
  ggsave(paste("g1_comp_vioPlt_", feature, ".pdf", sep = ""), plot = p, width = 10, height = 6)
  
    print(p)
}
```

# Compare the S-phase distribution of each feature in the EV and KRAS cells
```{r eval=FALSE, include=FALSE}
# Loop through each feature
for (feature in featureName) {
  # Combine the current feature column from each data frame into one data frame
  combined_df <- do.call(rbind, lapply(seq_along(s_data_frames), function(i) {
    df <- s_data_frames[[i]][, feature, drop = FALSE] # Select the current feature column
    df$SampleName <- sampleName[i] # Add the sample name
    df$FeatureName <- feature # Add the feature name
    return(df)
  }))
  
  # Rename the feature column to a generic name for plotting
  names(combined_df)[1] <- "Signal"
  
  # Create the violin plot
  p <- ggplot(combined_df, aes(x = SampleName, y = Signal, fill = SampleName)) +
    geom_violin(trim = FALSE) +
    scale_fill_brewer(palette = "Set3") + # Use a color palette for the fill
    labs(title = paste("S-phase violin Plot for", feature)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x labels for readability
  
  # Save the plot to the working directory with a unique name for each feature
  #ggsave(paste("g1_comp_vioPlt_", feature, ".pdf", sep = ""), plot = p, width = 10, height = 6)
  
    print(p)
}
```
# Compare the G1-phase distribution of each feature in the EV and KRAS cells
```{r}
# Loop through each feature
for (feature in featureName) {
  # Combine the current feature column from each data frame into one data frame
  combined_df <- do.call(rbind, lapply(seq_along(g1_data_frames), function(i) {
    df <- g2_data_frames[[i]][, feature, drop = FALSE] # Select the current feature column
    df$SampleName <- sampleName[i] # Add the sample name
    df$FeatureName <- feature # Add the feature name
    return(df)
  }))
  
  # Rename the feature column to a generic name for plotting
  names(combined_df)[1] <- "Signal"
  
  # Perform one-way ANOVA
  res.aov <- aov(Signal ~ SampleName, data = combined_df)
 print(sampleName)
  # Perform multiple comparison test
  res.tukey <- TukeyHSD(res.aov)
  colnames(res.tukey$SampleName) <- c('comparison', 'lwr', 'upr', 'padj')
  
  # Create the box and whisker plot
  p <- ggplot(combined_df, aes(x = SampleName, y = Signal, fill = SampleName)) +
    geom_boxplot() +
    scale_fill_brewer(palette = "Set3") + # Use a color palette for the fill
    labs(title = paste("G1-phase Box and Whisker Plot for", feature)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x labels for readability
  
  # Save the plot to the working directory with a unique name for each feature
  ggsave(paste("g1_comp_boxPlt_", feature, ".pdf", sep = ""), plot = p, width = 10, height = 6)
  
#  # Annotate the plot with the results of the multiple comparison test
#  p <- p + geom_text(data = as.data.frame(res.tukey$SampleName), aes(x = comparison, y = lwr, label = sprintf("p = #%.3f", padj)), vjust = -1.5)
  
  # Print the plot
  print(p)
}
```

# Compare the S-phase distribution of each feature in the EV and KRAS cells
```{r}
# Loop through each feature
for (feature in featureName) {
  # Combine the current feature column from each data frame into one data frame
  combined_df <- do.call(rbind, lapply(seq_along(s_data_frames), function(i) {
    df <- g2_data_frames[[i]][, feature, drop = FALSE] # Select the current feature column
    df$SampleName <- sampleName[i] # Add the sample name
    df$FeatureName <- feature # Add the feature name
    return(df)
  }))
  
  # Rename the feature column to a generic name for plotting
  names(combined_df)[1] <- "Signal"
  
  # Perform one-way ANOVA
  res.aov <- aov(Signal ~ SampleName, data = combined_df)
 print(sampleName)
  # Perform multiple comparison test
  res.tukey <- TukeyHSD(res.aov)
  colnames(res.tukey$SampleName) <- c('comparison', 'lwr', 'upr', 'padj')
  
  # Create the box and whisker plot
  p <- ggplot(combined_df, aes(x = SampleName, y = Signal, fill = SampleName)) +
    geom_boxplot() +
    scale_fill_brewer(palette = "Set3") + # Use a color palette for the fill
    labs(title = paste("S-phase Box and Whisker Plot for", feature)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x labels for readability
  
  # Save the plot to the working directory with a unique name for each feature
  ggsave(paste("s_comp_boxPlt_", feature, ".pdf", sep = ""), plot = p, width = 10, height = 6)
  
#  # Annotate the plot with the results of the multiple comparison test
#  p <- p + geom_text(data = as.data.frame(res.tukey$SampleName), aes(x = comparison, y = lwr, label = sprintf("p = #%.3f", padj)), vjust = -1.5)
  
  # Print the plot
  print(p)
}
```

# Compare the G2/M-phase distribution of each feature in the EV and KRAS cells
```{r}
# Loop through each feature
for (feature in featureName) {
  # Combine the current feature column from each data frame into one data frame
  combined_df <- do.call(rbind, lapply(seq_along(g2_data_frames), function(i) {
    df <- g2_data_frames[[i]][, feature, drop = FALSE] # Select the current feature column
    df$SampleName <- sampleName[i] # Add the sample name
    df$FeatureName <- feature # Add the feature name
    return(df)
  }))
  
  # Rename the feature column to a generic name for plotting
  names(combined_df)[1] <- "Signal"
  
  # Perform one-way ANOVA
  res.aov <- aov(Signal ~ SampleName, data = combined_df)
 print(sampleName)
  # Perform multiple comparison test
  res.tukey <- TukeyHSD(res.aov)
  colnames(res.tukey$SampleName) <- c('comparison', 'lwr', 'upr', 'padj')
  
  # Create the box and whisker plot
  p <- ggplot(combined_df, aes(x = SampleName, y = Signal, fill = SampleName)) +
    geom_boxplot() +
    scale_fill_brewer(palette = "Set3") + # Use a color palette for the fill
    labs(title = paste("G2M-phase Box and Whisker Plot for", feature)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x labels for readability
  
  # Save the plot to the working directory with a unique name for each feature
  ggsave(paste("g2_comp_boxPlt_", feature, ".pdf", sep = ""), plot = p, width = 10, height = 6)
  
#  # Annotate the plot with the results of the multiple comparison test
#  p <- p + geom_text(data = as.data.frame(res.tukey$SampleName), aes(x = comparison, y = lwr, label = sprintf("p = #%.3f", padj)), vjust = -1.5)
  
  # Print the plot
  print(p)
}
```

## Trying ridgeline plot to see if the plots look better
```{r eval=FALSE, include=FALSE}
for (feature in featureName) {
  # Combine the current feature column from each data frame into one data frame
  combined_df <- do.call(rbind, lapply(seq_along(data_frames), function(i) {
    df <- data_frames[[i]][, feature, drop = FALSE] # Select the current feature column
    df$SampleName <- sampleName[i] # Add the sample name
    df$FeatureName <- feature # Add the feature name
    return(df)
  }))
  
  # Rename the feature column to a generic name for plotting
  names(combined_df)[1] <- "Measurement"
 
    # Create the ridge line plot
    p <- ggplot(combined_df, aes(x = Measurement, y = SampleName, fill = SampleName)) +
      geom_density_ridges() +
      scale_fill_brewer(palette = "Set3") + # Use a color palette for the fill
      labs(title = paste("Ridge Line Plot for", feature)) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) # Rotate x labels for readability
    
    # Save the plot to the working directory with a unique name for each feature
    ggsave(paste("comp_rdgLine_", feature, ".pdf", sep = ""), plot = p, width = 10, height = 6)
    print(p)
}
```

## Density plots
```{r eval=FALSE, include=FALSE}
for (feature in featureName) {
    # Combine the current feature column from each data frame into one data frame
    combined_df <- do.call(rbind, lapply(seq_along(data_frames), function(i) {
        df <- data_frames[[i]][, feature, drop = FALSE] # Select the current feature column
        df$SampleName <- sampleName[i] # Add the sample name
        df$FeatureName <- feature # Add the feature name
        return(df)
    }))
    
    # Rename the feature column to a generic name for plotting
    names(combined_df)[1] <- "Measurement"
    
    # Create the density plot
    p <- ggplot(combined_df, aes(x = Measurement, fill = SampleName)) +
        geom_density(alpha = 0.65) +
        scale_fill_brewer(palette = "Set3") + # Use a color palette for the fill
        labs(title = paste("Density Plot for", feature)) +
        theme_minimal()
    
    # Print the plot
    print(p)
    
    # Save the plot to the working directory with a unique name for each feature
    ggsave(paste("comp_dnsty_", feature, ".pdf", sep = ""), plot = p, width = 10, height = 6)
}
```

## Pair-wise density plots to show the difference within the same cell line
```{r Pair-wise plots, eval=FALSE, include=FALSE}
# Function to create mirrored density plots
pairwise_density_plot <- function(df1, df2, n_df1, n_df2, f1, f2, col) {
    # Create a new plot
    p <- ggplot() +
        geom_density(data = df1, aes(x = .data[[col]], fill = f1), alpha = 0.7) +
        geom_density(data = df2, aes(x = .data[[col]], fill = f2), linetype = "dashed", alpha = 0.7) +
        scale_fill_brewer(palette = "Set3") +  
      #scale_fill_manual(values = c(n_df1 = "lightblue", n_df2 = "red")) +
        labs(title = paste("Pair-wise Density Plot for", col),
             x = col,
             y = "Density") +
        theme_minimal()
    
    return(p)
}

# Find the indices of EV and EV200 data frames
ev_index <- match("EV", sampleName)
ev200_index <- match("EV200", sampleName)

# Create mirrored density plots for EV vs. EV200
for (feature in featureName) {
    plot <- pairwise_density_plot(data_frames[[ev_index]], data_frames[[ev200_index]],
                                       names(data_frames[1])[1], names(data_frames[2])[1],
                                       sampleName[ev_index], sampleName[ev200_index], feature)

    # Save the plot to the working directory with a unique name for each feature
    ggsave(paste("EV_pair_dnsty_", feature, ".pdf", sep = ""), plot = plot, width = 10, height = 6)
}

# Find the indices of KRAS and KRAS200 data frames
kras_index <- match("KRAS", sampleName)
kras200_index <- match("KRAS200", sampleName)

for (feature in featureName) {
    plot <- pairwise_density_plot(data_frames[[kras_index]], data_frames[[kras200_index]],
                                       names(data_frames[3]), names(data_frames[4]),
                                       sampleName[kras_index],sampleName[kras200_index],feature)

    # Save the plot to the working directory with a unique name for each feature
    ggsave(paste("KRAS_pair_dnsty_", feature, ".pdf", sep = ""), plot = plot, width = 10, height = 6)
    print(plot)
}
```

## Output filtered data-sets to plot in Jupyter notebook
```{r Write filtered data, eval=FALSE, include=FALSE}
# Use lapply to apply the function to each element in the list
lapply(names(data_frames), function(x) {
    filename <- paste0("R_",x, "_filtered.csv")
    counter <- 0
    while(file.exists(filename)) {
        counter <- counter + 1
        filename <- paste0(x, "_filtered", counter, ".csv")
    }
    write.csv(data_frames[[x]], filename, row.names = FALSE)
})
```